Evaluating Informative Hypotheses with Equality and Inequality Constraints:\\ A Tutorial Using the Bayes Factor via the Encompassing Prior Approach

### 2022-10-09 Documents/Paper/paper.Rnw
revision foor resubmission

\abstract{When conducting a study, researchers usually have expectations based on hypotheses or theoretical perspectives they want to evaluate. Equality and inequality constraints on the model parameters are used to formalize researchers' expectations or theoretical perspectives into the so-called informative hypotheses. However, traditional statistical approaches, such as the Null Hypothesis Significance Testing (NHST) or the model comparison using information criteria (e.g., AIC and BIC), are unsuitable for testing complex informative hypotheses. An alternative approach is to use the Bayes factor. In particular, the Bayes factor based on the encompassing prior approach allows researchers to easily evaluate complex informative hypotheses in a wide range of statistical models (e.g., generalized linear). This paper provides a detailed introduction to the Bayes factor with encompassing prior. First, all steps and elements involved in the formalization of  informative hypotheses and the computation of the Bayes factor with encompassing prior are described. Next, we apply this method to a real case scenario, considering the attachment theory. Specifically, we analyzed the relative influence of maternal and paternal attachment on children's social-emotional development by comparing the various theoretical perspectives debated in the literature.

\vspace{2em}
\begin{center}
\textbf{Translational Abstract}
\end{center}
\vspace{.5em}

The evaluation of hypotheses is one of the main goals of empirical research. In fact, when conducting a study, researchers usually want to evaluate the plausibility of specific hypotheses according to the collected data. These hypotheses can be formalized in accordance with researchers' expectations or theoretical perspectives into the so-called informative hypotheses. However, traditional statistical approaches, such as the Null Hypothesis Significance Testing (NHST) or the model comparison using information criteria (e.g., AIC and BIC), are not suitable for evaluating the plausibility of complex informative hypotheses. Alternatively, the Bayes factor via the encompassing prior approach allows researchers to easily evaluate the plausibility of complex informative hypotheses in a wide range of statistical models. This paper provides a detailed introduction to the Bayes factor with encompassing prior. First, we describe all steps and elements involved in the formalization of  informative hypotheses and the computation of the Bayes factor. Next, we apply this method to a real case scenario, considering the attachment theory. Specifically, we analyzed the relative influence of maternal and paternal attachment on children's social-emotional development by comparing the various theoretical perspectives debated in the literature. This approach is of particular interest in applied research, as it allows researchers to formalize hypotheses with great flexibility and to directly evaluate which hypothesis is the most supported by the data.
}

@@

Model comparison is a different approach that allows researchers to compare multiple hypotheses and identify the most supported by the data \parencite{mcelreathStatisticalRethinkingBayesian2020}. Researchers first formalize hypotheses as statistical models according to their expectations or theoretical perspectives. Therefore, it is possible to determine which is the most supported model among those considered according to the data. To do so, a popular approach is to use information criteria, such as the AIC or BIC criteria, that estimate a model's ability to predict new data \parencite{wagenmakersAICModelSelection2004,akaike1973a, schwarzEstimatingDimensionModel1978}. Model comparison has several advantages; in particular, it allows researchers to directly compute each model's relative plausibility given the data and the set of models considered. However, model comparison using information criteria is not appropriate to evaluate hypotheses that include complex parameter constraints reflecting researchers' expectations. In fact, information criteria evaluate models' complexity according to the number of parameters, but they do not take into account possible order constraints on the parameters.

An alternative criterium to evaluate research hypotheses in a model comparison is the Bayes factor. During the last 25 years, interest in the Bayes factor has increased, and its use has been proposed as the solution to the NHST's critical issues \parencite{heckReviewApplicationsBayes2020, mulderEditorsIntroductionSpecial2016}. Although it has its own limitations and it does not solve the fundamental issues of the misuse of statistical techniques \parencite{schadWorkflowTechniquesRobust2021, gelmanBayesianDataAnalysis2013}, the Bayes factor offers some clear advantages. In particular, it allows us to compare hypotheses obtaining a relative index of evidence, like the information criteria, but it also allows us to easily compare complex research hypotheses. In fact, researchers can formalize so-called \textit{“informative hypotheses”}, hypotheses containing information about the ordering of the model parameters,  according to their expectations or theoretical perspectives, using equality and inequality constraints \parencite{vandeschootIntroductionBayesianModel2011}. Researchers can then compare these hypotheses against each other, using the Bayes factor.

The evaluation of informative hypotheses is of particular interest to researchers, as it allows them to assess specific complex expectations and theoretical perspectives directly. In the literature, a particular approach allowing researchers to easily compute the Bayes factor with complex informative hypotheses has received increasing attention: \textit{“the Bayes factor with encompassing prior”}. \textcite{vandeschootIntroductionBayesianModel2011} presented a general introduction to informative hypothesis testing using the Bayes factor with the encompassing prior approach whereas \textcite{hoijtinkInformativeHypothesesTheory2012} offered a more detailed description of this method's development. In other studies, however, researchers considered the application of this approach with specific statistical models, such as mixed-effect models \parencite{katoBayesianApproachInequality2006}, evaluation of correlation coefficients \parencite{mulderBayesFactorsTesting2016}, and multinomial models \parencite{heckMultinomialModelsLinear2019}. In addition, \textcite{guBayesianEvaluationInequality2014} proposed a general approximate procedure to evaluate inequality constraints in a wide range of statistical models.

@@

Only recently, \textcite{guApproximatedAdjustedFractional2018} introduced an approximate procedure to evaluate equality and inequality constraints, using the Bayes factor with the encompassing prior approach in a wide range of statistical models (i.e., generalized linear mixed models and structural equation models). Subsequently, \textcite{mulderBayesFactorTesting2019} extended this approach to generalized multivariate probit models, and, finally, \textcite{mulderSimpleBayesianTesting2019} proposed an accurate procedure (i.e., not based on an approximation) that allows for testing of informative hypotheses with equality and inequality constraints in linear regression models.

The development and implementation of this approach are of particular interest because although it has limits, it allows researchers to directly evaluate their expectations and hypotheses. The available literature, however, is rather technical. These articles' complexity makes it difficult for researchers not familiar with this approach to clearly understand all the steps involved. On the other hand, articles offering a general introduction to the Bayes factor do not provide enough details to allow readers to autonomously apply this method to their problems; instead, they usually rely on ad hoc solutions implemented in some statistical software. For example, \textcite{hoijtinkTutorialTestingHypotheses2019} offered an excellent introduction to the Bayes factor for the evaluation of informative hypotheses using the R package \texttt{bain}. However, it is limited to the context of analysis-of-variance models, and it is not possible to extend its use to other cases.

The aim of this paper, therefore, is to offer a clear and detailed description of the Bayes factor with the encompassing prior approach to allow other researchers to apply this approach in their studies. In particular, we refer to the approximated method \textcite{guApproximatedAdjustedFractional2018} proposed, as it applies to a wider range of conditions than the more accurate approach \textcite{mulderSimpleBayesianTesting2019} introduced. In particular, researchers can use the proposed method in the case of generalized linear mixed models and structural equation models, two very general families of statistical models. Moreover, by providing an exhaustive description of all of this approach's steps, researchers do not have to rely on already available solutions; instead, they can implement this approach according to their specific needs. The paper is organized as follows. First, we introduce the method, providing a detailed description of all steps and elements involved in the formalization of informative hypotheses and Bayes factor computation. We then present an application of the method to real data to discuss the typical issues encountered in real complex scenarios.

@@

The evaluation of informative hypotheses with equality and/or inequality constraints involves several steps and elements. In this section, first, we describe how to formulate informative hypotheses. We then introduce the Bayes factor, considering the encompassing prior approach based on the approximated method \textcite{guApproximatedAdjustedFractional2018} proposed.

@@

The first term is the ratio between the conditional posterior probability and the conditional prior probability that the inequality constraints hold under the unconstrained model, $H_u$, given the equality constraints. The second term is the ratio between marginal posterior density and the marginal prior density of the equality constraints under $H_u$ \parencite[the well-known Savage–Dickey density ratio; ][]{dickeyWeightedLikelihoodRatio1971, wetzelsEncompassingPriorGeneralization2010}. In particular, the four elements can be interpreted as:
\begin{itemize}
  \item{The \textbf{conditional posterior probability}, $Pr(R_I\theta > r_I | R_E\theta =r_E, Y, H_u)$, is a measure of the fit of the inequality constraints of $H_i$ under $H_u$.}
  \item{The \textbf{conditional prior probability}, $\pi(R_I\theta > r_I | R_E\theta =r_E, H_u)$, is a measure of the complexity of the inequality constraints of $H_i$ under $H_u$.}
  \item{The \textbf{marginal posterior density}, $Pr(R_E\theta =r_E| Y, H_u)$, is a measure of the fit of the equality constraints of $H_i$ under $H_u$.}
  \item{The \textbf{marginal prior density}, $\pi(R_E\theta =r_E| H_u)$, is a measure of the complexity of the equality constraints of $H_i$ under $H_u$.}
\end{itemize}
Summarizing at the denominators, we have measures of the \textbf{complexity} of the inequality and equality constraints of the informative hypothesis $H_i$; at the numerators, instead, we have measures of the \textbf{fit} of the data to the inequality and equality constraints of the informative hypothesis $H_i$. Therefore, if the hypothesis $H_i$, although applying constraints to the parametric space (less complexity), can still provide an accurate description of the data, the Bayes factor will favor $H_i$. On the contrary, if the support of the data is poor, the Bayes factor will favor the unconstrained model, $H_u$.

@@

@@

We used the following notation to formalize the hypotheses. $M$ and $F$ are used to indicate attachment toward the mother and the father, respectively. Specific attachment styles are specified using subscript: $S$ indicates secure attachment, $Ax$ anxious attachment, $Av$ avoidant attachment, and $F$ fearful attachment. For example, $F_{Av}$ represents children with avoidant attachment toward their fathers.

Note that when we do not expect interaction between mother and father attachment, we can consider the roles of the two parents separately. Whereas, if an interaction is expected, it is necessary to take into account the unique combination of mother and father attachment. Thus, for example, we use $M_SF_{Ax}$ to indicate children with secure attachment toward their mothers and anxious attachment toward their fathers. Moreover, $*$ subscript is used to indicate any attachment style and a set of subscripts is used to indicate “\textit{one among}”. For example, $M_SF_{Ax;Av}$ represents children with secure attachment toward their mothers and anxious or avoidant attachment toward their fathers.

@@

To interpret these patterns, remember that order constraints are insensitive to the distribution specification as long as the distribution is symmetric and centered on the constraint focal point. On the contrary, equality constraints are highly affected by the prior definition. As the prior gets more diffuse, the density value at zero decreases as well and, even for posterior distributions centered far from zero, the densities ratio at zero will favor the posterior. Thus, as the prior becomes more diffuse, the Bayes factor will increasingly favor the hypotheses with equality constraints. For tighter priors, the Bayes factor will strongly penalize hypotheses with equality constraints if these are not correct (see Figure~\ref{fig:plot-sensitivity-prior}).

>>>>

\abstract{When conducting a study, researchers usually have expectations based on hypotheses or theoretical perspectives they want to evaluate. Equality and inequality constraints on the model parameters are used to formalize researchers' expectations or theoretical perspectives into the so-called informative hypotheses. However, traditional statistical approaches, such as the Null Hypothesis Significance Testing (NHST) or the model comparison using information criteria (e.g., AIC and BIC), are unsuitable for testing complex informative hypotheses. An alternative approach is to use the Bayes factor. In particular, the Bayes factor based on the encompassing prior approach allows researchers to easily evaluate complex informative hypotheses in a wide range of statistical models (e.g., generalized linear). This paper provides a detailed introduction to the Bayes factor with encompassing prior. First, all steps and elements involved in the formalization of  informative hypotheses and the computation of the Bayes factor using the encompassing prior approach are described. Next, we apply this method to a real-case scenario, considering the attachment theory. Specifically, we analyzed the relative influence of maternal and paternal attachment on children's social-emotional development by comparing the various theoretical perspectives debated in the literature.

% \vspace{2em}
% \begin{center}
% \textbf{Translational Abstract}
% \end{center}
% \vspace{.5em}
%
% The evaluation of hypotheses is one of the main goals of empirical research. In fact, when conducting a study, researchers usually want to evaluate the plausibility of specific hypotheses according to the collected data. These hypotheses can be formalized in accordance with researchers' expectations or theoretical perspectives into the so-called informative hypotheses. However, traditional statistical approaches, such as the Null Hypothesis Significance Testing (NHST) or the model comparison using information criteria (e.g., AIC and BIC), are not suitable for evaluating the plausibility of complex informative hypotheses. Alternatively, the Bayes factor via the encompassing prior approach allows researchers to easily evaluate the plausibility of complex informative hypotheses in a wide range of statistical models. This paper provides a detailed introduction to the Bayes factor with encompassing prior. First, we describe all steps and elements involved in the formalization of  informative hypotheses and the computation of the Bayes factor. Next, we apply this method to a real case scenario, considering the attachment theory. Specifically, we analyzed the relative influence of maternal and paternal attachment on children's social-emotional development by comparing the various theoretical perspectives debated in the literature. This approach is of particular interest in applied research, as it allows researchers to formalize hypotheses with great flexibility and to directly evaluate which hypothesis is the most supported by the data.
}

@@

Model comparison is a different approach that allows researchers to compare multiple hypotheses and identify the one most supported by the data \parencite{mcelreathStatisticalRethinkingBayesian2020}. Researchers first formalize hypotheses as statistical models according to their expectations or theoretical perspectives. Therefore, it is possible to determine which is the most supported model among those considered according to the data. To do so, a popular approach is to use information criteria, such as the AIC or BIC criteria, that estimate a model's ability to predict new data \parencite{wagenmakersAICModelSelection2004,akaike1973a, schwarzEstimatingDimensionModel1978}. Model comparison has several advantages; in particular, it allows researchers to directly compute each model's relative plausibility given the data and the set of models considered. However, model comparison using traditional information criteria (i.e., AIC and BIC) is not appropriate to evaluate hypotheses that include complex parameter constraints reflecting researchers' expectations. In fact, traditional information criteria penalize models' complexity according to the number of parameters, but they do not take into account possible order constraints on the parameters. Modifications of the AIC have been proposed to account for parameters order restrictions, leading to the order-restricted information criterion \parencite[ORIC;][]{anrakuInformationCriterionParameters1999} and its recent generalization into the so-called generalized order-restricted information criterion approximation \parencite[GORICA;][]{altinisikEvaluationInequalityConstrained2021, kuiperAkaiketypeInformationCriterion2011, kuiperGeneralizationOrderrestrictedInformation2012}. The GORICA allows to evaluate (in)equality constrained hypotheses under a very broad range of statistical models \parencite{altinisikEvaluationInequalityConstrained2021, kuiperAICtypeTheoryBasedModel2022}, including generalized linear models (GLMs), generalized linear mixed models (GLMMs), and structural equation models (SEMs). Consider also \textcite{mulderBICExtensionsOrderconstrained2022} for an extension of the BIC for evaluating order-constrained models.

An alternative criterium to evaluate research hypotheses in a model comparison is the Bayes factor. During the last 25 years, interest in the Bayes factor has increased, and its use has been proposed as the solution to the NHST's critical issues \parencite{heckReviewApplicationsBayes2020, mulderEditorsIntroductionSpecial2016}. Although it has its own limitations and it does not solve the fundamental issues of the misuse of statistical techniques \parencite{schadWorkflowTechniquesRobust2021, gelmanBayesianDataAnalysis2013}, the Bayes factor offers some clear advantages. In particular, it allows us to compare hypotheses obtaining a relative index of evidence, like the information criteria, but it also allows us to easily compare complex research hypotheses. In fact, researchers can formalize the so-called \textit{“informative hypotheses”}, hypotheses containing information about the ordering of the model parameters,  according to their expectations or theoretical perspectives, using equality and inequality constraints \parencite{vandeschootIntroductionBayesianModel2011}. Researchers can then compare these hypotheses against each other, using the Bayes factor.

The evaluation of informative hypotheses is of particular interest to researchers, as it allows them to assess specific complex expectations and theoretical perspectives directly. In the literature, a particular approach allowing researchers to easily compute the Bayes factor with complex informative hypotheses has received increasing attention: \textit{“the Bayes factor with encompassing prior”}. \textcite{vandeschootIntroductionBayesianModel2011} presented a general introduction to informative hypothesis testing using the Bayes factor with the encompassing prior approach whereas \textcite{hoijtinkInformativeHypothesesTheory2012, klugkistBayesianModelSelection2005, klugkistBayesFactorInequality2007} offered a more detailed description of this method. In other studies, researchers considered the application of this approach with specific statistical models, such as analysis of variance models \parencite{klugkistInequalityConstrainedAnalysis2005a}, mixed-effect models \parencite{katoBayesianApproachInequality2006}, evaluation of correlation coefficients \parencite{mulderBayesFactorsTesting2016}, and multinomial models \parencite{heckMultinomialModelsLinear2019}. In addition, \textcite{guBayesianEvaluationInequality2014} proposed a general approximate procedure to evaluate inequality constraints in a wide range of statistical models.

@@

Only recently, \textcite{guApproximatedAdjustedFractional2018} introduced an approximate procedure to evaluate equality and inequality constraints, using the Bayes factor with the encompassing prior approach in a wide range of statistical models (i.e., GLMs, GLMMs, and SEMs). Subsequently, \textcite{mulderBayesFactorTesting2019} extended this approach to generalized multivariate probit models, and, finally, \textcite{mulderSimpleBayesianTesting2019} proposed an accurate procedure (i.e., not based on an approximation) that allows for testing of informative hypotheses with equality and inequality constraints in linear regression models.

The development and implementation of these methods are of particular interest because although they have some limits, they allow researchers to directly evaluate their expectations and hypotheses. The available literature, however, is rather technical. The complexity of these articles makes it difficult for researchers not familiar with this approach to clearly understand all the steps involved. On the other hand, articles offering a general introduction to the Bayes factor do not provide enough details to allow readers to autonomously apply this method to their problems; instead, they usually rely on ad hoc solutions implemented in some statistical software. For example, \textcite{hoijtinkTutorialTestingHypotheses2019} offered an excellent introduction to the Bayes factor for the testing hypotheses discussing all the main aspects. However, the description of the evaluation of competing informative hypotheses is limited and no specific details are provided regarding the adopted encompassing prior approach. Moreover, using ad-hoc solutions that automate the analysis, such as the functions available in the R packages \texttt{bain} \parencite{guBainBayesFactors2021} and \texttt{BFpack} \parencite{mulderBFpackFlexibleBayes2021}, prevent the users from fully understanding all the steps and limits entailed in this approach.

The aim of this paper, therefore, is to offer a clear and detailed description of the Bayes factor with the encompassing prior approach to allow other researchers to consciously apply this approach in their studies. In particular, we refer to the approximated method \textcite{guApproximatedAdjustedFractional2018} proposed, as it applies to a wider range of conditions than the more accurate approach \textcite{mulderSimpleBayesianTesting2019} introduced. In particular, researchers can use the proposed method in the case of generalized linear mixed models and structural equation models, two very general families of statistical models. Moreover, by providing an exhaustive description of all of the steps of this approach, researchers do not have to rely on already available solutions; instead, they can implement this approach according to their specific needs, aware of its strengths and limits. The paper is organized as follows. First, we introduce the method, providing a detailed description of all steps and elements involved in the formalization of informative hypotheses and Bayes factor computation. We then present an application of the method to real data to discuss the typical issues encountered in real complex scenarios.

@@

The evaluation of informative hypotheses with equality and/or inequality constraints involves several steps and elements. In this section, first, we describe how to formulate informative hypotheses. We then introduce the Bayes factor, considering the encompassing prior approach based on the approximated method proposed by \textcite{guApproximatedAdjustedFractional2018}.

@@

The first term is the ratio between the conditional posterior probability and the conditional prior probability that the inequality constraints hold under the unconstrained model, $H_u$, given the equality constraints. The second term is the ratio between marginal posterior density and the marginal prior density of the equality constraints under $H_u$ \parencite[the well-known Savage–Dickey density ratio; Equation (6) can be seen as an extension of the Savage-Dickey method, where the density ratio of the equality constraints is multiplied by the probability ratio of the posterior and prior.][]{dickeyWeightedLikelihoodRatio1971, wetzelsEncompassingPriorGeneralization2010}. In particular, the four elements can be interpreted as:
\begin{itemize}
  \item{The \textbf{conditional posterior probability}, $Pr(R_I\theta > r_I | R_E\theta =r_E, Y, H_u)$, is a measure of the fit of the inequality constraints of $H_i$ under $H_u$.}
  \item{The \textbf{conditional prior probability}, $\pi(R_I\theta > r_I | R_E\theta =r_E, H_u)$, is a measure of the complexity of the inequality constraints of $H_i$ under $H_u$.}
  \item{The \textbf{marginal posterior density}, $Pr(R_E\theta =r_E| Y, H_u)$, is a measure of the fit of the equality constraints of $H_i$ under $H_u$.}
  \item{The \textbf{marginal prior density}, $\pi(R_E\theta =r_E| H_u)$, is a measure of the complexity of the equality constraints of $H_i$ under $H_u$.}
\end{itemize}

Summarizing at the denominators, we have measures of the \textbf{complexity} of the inequality and equality constraints of the informative hypothesis $H_i$; at the numerators, instead, we have measures of the \textbf{fit} of the data to the inequality and equality constraints of the informative hypothesis $H_i$. Where the \textit{fit} of an hypothesis indicates its ability to provide a good description of data at hand and the \textit{complexity} of an hypothesis refers to its level of specificity and parsimony (i.e., simpler hypothesis are specific and specific and parsimonious). Therefore, if the hypothesis $H_i$, although applying constraints to the parametric space (less complexity), can still provide an accurate description of the data, the Bayes factor will favor $H_i$. On the contrary, if the support of the data is poor, the Bayes factor will favor the unconstrained model, $H_u$.

@@

Note that, prior specification is relevant only for the parameters of interest (i.e., the parameters involved in the constraints). Nuisance parameters (i.e., the parameters not involved in the constraints) are integrated out in the Bayes factor. Thus, the results will be insensitive to the actual choice of the prior for the nuisance parameters as long as it is vague enough.

@@

We used the following notation to formalize the hypotheses. $M$ and $F$ are used to indicate attachment toward the mother and the father, respectively. Specific attachment styles are specified using subscript: $S$ indicates secure attachment, $Ax$ anxious attachment, $Av$ avoidant attachment, and $F$ fearful attachment. For example, $F_{Av}$ represents children with avoidant attachment toward their fathers. Moreover, $*$ subscript is used to indicate any attachment style and a set of subscripts is used to indicate “\textit{one among}”. For example, $F_{*}$ indicates children with any type of attachment toward their fathers and $F_{Ax;Av}$ represents children with anxious or avoidant attachment toward their fathers.

Note that when we do not expect interaction between mother and father attachment, we can consider the roles of the two parents separately. Whereas, if an interaction is expected, it is necessary to take into account the unique combination of mother and father attachment. Thus, for example, we use $M_SF_{Ax}$ to indicate children with secure attachment toward their mothers and anxious attachment toward their fathers.

@@

To interpret these patterns, remember that the bayes factor is insensitive to the prior distribution specification for the parameters with order constraints as long as the distribution is symmetric and centered on the constraint focal point. On the contrary, in the case of equality constraints, the bayes factor is highly affected by the prior definition. As the prior gets more diffuse, the density value at zero decreases as well and, even for posterior distributions centered far from zero, the densities ratio at zero will favor the posterior. Thus, as the prior becomes more diffuse, the Bayes factor will increasingly favor the hypotheses with equality constraints. For tighter priors, the Bayes factor will strongly penalize hypotheses with equality constraints if these are not correct (see Figure~\ref{fig:plot-sensitivity-prior}).
